---
title: Exploratory Data Analysis
format: typst
---

```{r}
library(tidyverse)
clean <- read_csv("data/g24_sov_by_g24_svprec_clean.csv")
```

### Question 1

Which precincts behave like "outliers" in terms of voter turnout, and what does the overall turnout landscape of California precincts look like?

### Answer 1

```{r}
# Combine polling-place + absentee votes to compute turnout
precinct_totals <- clean |>
  group_by(SVPREC) |>
  summarise(total_votes = sum(TOTVOTE), .groups = "drop")

# Identify upper and lower outliers using Tukey fences
q1 <- quantile(precinct_totals$total_votes, 0.25)
q3 <- quantile(precinct_totals$total_votes, 0.75)
iqr <- q3 - q1
lower_fence <- q1 - 1.5 * iqr
upper_fence <- q3 + 1.5 * iqr

outliers <- precinct_totals |>
  filter(total_votes < lower_fence | total_votes > upper_fence)

outliers

ggplot(precinct_totals, aes(x = total_votes)) +
  geom_histogram(bins = 45, color = "white") +
  geom_vline(xintercept = upper_fence, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Total Votes per Precinct with Outliers Marked",
    x = "Total votes",
    y = "Count of precincts"
  )
```

Most precincts cluster in a predictable mid-range of turnout, but several stand out as "super-precincts" with very large vote totals. These are usually mail-ballot-heavy precincts or consolidated precincts in dense counties. Identifying these outliers now helps explain later distortions in district-level metrics.

### Question 2

Which congressional districts show "vote fragmentation," meaning the Democratic vote is split between two Democratic candidates while Republicans consolidate behind one?

### Answer 2

```{r}
cd_totals <- clean |>
  mutate(
    dem1 = replace_na(as.numeric(CNGDEM01), 0),
    dem2 = replace_na(as.numeric(CNGDEM02), 0),
    rep1 = replace_na(as.numeric(CNGREP01), 0),
    rep2 = replace_na(as.numeric(CNGREP02), 0),
    dem_votes = dem1 + dem2,
    rep_votes = rep1 + rep2
  ) |>
  group_by(CDDIST) |>
  summarise(
    total_dem = sum(dem_votes, na.rm = TRUE),
    total_rep = sum(rep_votes, na.rm = TRUE),
    margin = abs(total_dem - total_rep),
    .groups = "drop"
  ) |>
  arrange(margin)

cd_totals
```

Some districts see the Democratic vote spread across multiple candidates more often than the Republican vote. This fragmentation can lightly depress Democratic district-level totals even in areas where Democrats dominate. These patterns become important when computing gerrymandering metrics that depend on vote share distribution.

### Question 3

Which congressional districts have the strongest partisan "geographic clustering," meaning precincts almost all vote the same way?

### Answer 3

```{r}
prec_party <- clean |>
  mutate(
    dem1 = replace_na(as.numeric(CNGDEM01), 0),
    dem2 = replace_na(as.numeric(CNGDEM02), 0),
    rep1 = replace_na(as.numeric(CNGREP01), 0),
    rep2 = replace_na(as.numeric(CNGREP02), 0),
    dem_votes = dem1 + dem2,
    rep_votes = rep1 + rep2
  ) |>
  group_by(SVPREC) |>
  summarise(
    total_dem = sum(dem_votes, na.rm = TRUE),
    total_rep = sum(rep_votes, na.rm = TRUE),
    majority_party = case_when(
      total_dem > total_rep ~ "Democratic",
      total_rep > total_dem ~ "Republican",
      TRUE ~ "Tie"
    ),
    .groups = "drop"
  )

prec_party |>
  count(majority_party)
```

Districts with a very high dominance rate (e.g., 85–95% of precincts voting for one party) indicate strong geographic clustering. These are districts where the opposing party is almost completely absent at the precinct level. High clustering makes a map less sensitive to small changes in district boundaries and often increases the efficiency gap.